Fitness tracking reveals task-specific associations between memory, mental health, and physical activity

### 2022-07-18 paper/main.tex
use mac-style bib path
final round of edits, recompiled diffs

\paragraph{Languages.}
All participants reported that they were fluent in either 1 and 2
languages (25\textsuperscript{th} percentile: 1;
50\textsuperscript{th} percentile: 1; 75\textsuperscript{th}
percentile: 1), and that they were ``familiar'' with between 1 and 11
languages (25\textsuperscript{th} percentile: 1;
50\textsuperscript{th} percentile: 2; 75\textsuperscript{th}
percentile: 3).

@@

After studying all 10 flashcards, participants were given a multiple
choice memory test where they were shown a series of novel
photographs, each depicting one of the 10 words they had learned.
They were asked to select which (of 4 unique options) Gaelic word went
with the given picture.  The 3 incorrect options were selected at
random (with replacement across trials), and the order in which the
choices appeared to the participant were also randomized.  Each of the
10 words they had learned were tested exactly once.

@@

\section*{Results}
Before testing our main hypotheses, we examined the behavioral data
from each of four memory tasks: a random word list learning ``free
recall'' task (Fig.~\ref{fig:tasks}); a naturalistic recall task whereby participants watched
a short video and then recounted the narrative; a foreign language
``flashcards'' task; and a spatial learning task.  Each of the first
three tasks (free recall, naturalistic recall, and the flashcards
task) included both an immediate (short term) memory test and a
delayed (long term) memory test.  The spatial learning task included
only an immediate test. Participants in all four tasks exhibited
general trends and tendencies that have been previously reported in
prior work.  We were also interested in characterizing the variability
in task performance across participants.  For example, if all
participants exhibited near-identical behaviors or performance on a
given task, we would be unable to identify how memory performance on
that task varied with mental health or physical activity.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/combined_correlations_main}
\caption{\textbf{Summaries of correlations between behavioral,
    fitness, and mental health measures.}  The reported values in the
  tables reflect correlations between each behavioral measure and
  mental health measure.  Only statistically reliable correlations
  ($p < 0.05$, corrected) are displayed.  \textbf{a. Correlations
    between behavioral and mental health measures.}  We adjusted each
  task's behavioral measure(s) such that more positive values reflect
  better performance on the given task.  We used participants' mean
  recall accuracy to characterize performance on the free recall and
  foreign language flashcards tasks, and mean precision to
  characterize performance on the naturalistic recall tasks.  We
  characterized performance on the spatial learning task using the
  (inverted and normalized) intercepts and slopes of linear
  regressions on mean estimation errors as a function of the number of
  studied shapes (also see Figs.~\ref{fig:immediate_behavior},
  \ref{fig:delayed_behavior},~\frDetail,~\natDetail,~\vocabDetail,
  and~\spatialDetail).  For each mental health measure, more positive
  values denote greater severity of the given measure.  Typical and
  current stress levels were measured by self report.  Mental health
  information was inferred using each participants' list of
  self-reported medications (see \textit{Methods}).  Positive
  correlations indicate that better performance on a given behavioral
  task is associated with more severe mental health phenotypes.
  \textbf{b. Correlations between fitness and mental health measures.}
  For each fitness measure, more positive values denote higher
  observed scores (i.e., higher resting heart rate, more minutes of
  activity or time spent in each heart rate zone, or greater heart
  rate variability).  The mental health measures in this panel were
  treated as in Panel a.  \textbf{c. Correlations between fitness and
    behavioral measures.}  Each measure reflected in this panel was
  treated as in Panels a and b.}
\label{fig:corr_summaries}
\end{figure}

We also found reliable correlations between participants' fitness and
demographic measures and their behaviors in different tasks
(Fig.~\ref{fig:corr_summaries}; for a complete report, see Fig.~\allCorr).
For example, recent low-to-moderate-intensity (``fat burn'')
cardiovascular activity was positively correlated with immediate
($r = 0.44, p = 0.001$) and delayed ($r = 0.38, p = 0.031$) recall
performance on the naturalistic memory task.  Recent sedentary
(``out-of-range'') cariovascular activity was negatively correlated
with performance on the spatial learning task
($r = -0.31, p = 0.042$), whereas recent high intensity (``peak'')
activity was positively correlated with performance on the spatial
learning task ($r = 0.34, p = 0.0002$).  Mental health indicators,
such as self-reported stress levels and medications were also
associated with differences in memory
(Figs.~\ref{fig:corr_summaries}a, \allCorr).  For example,
self-reported stress levels at the time of test were negatively
correlated with performance on the delayed memory test for the foreign
language flashcards task ($r = -0.29, p = 0.038$), whereas
participants who were medicated for anxiety and depression tended to
perform slightly (but reliably) \textit{better} on the immediate
memory test for the foreign language flashcards task
($r = 0.11, p < 0.0001$).  Mental health indicators were also
correlated with several fitness measures
(Fig.~\ref{fig:corr_summaries}c).  For example, participants with
higher resting heart rates were less likely to be hypothyroid ($r =
-0.33, p < 0.0001$).  Participants who engaged in more low-intensity
(``light'') activity tended to be less anxious and depressed ($r =
-0.12, p = 0.03$), whereas participants who engaged in more
high-intensity activity tended to report higher levels of current ($r
= 0.15, p = 0.027$) and typical ($r = 0.21, p = 0.012$) stress.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/weighted_timecourse_summary}
\caption{\textbf{Dynamics of physical activity varies with memory
    performance and mental health measures.}  \textbf{a. Daily step counts.}  Each
  timecourse is weighted by either performance on immediate recall tests
  (left panel) or on delayed recall tests (right panel).  The black
  (baseline) timecourses display the (unweighted) average across all
  participants.  \textbf{b. Daily duration (in minutes) of low-intensity physical
    activity.}  Timecourses are displayed in the same format and color
scheme as those in Panel A.  Analogous timecourses for additional
fitness-related measures may be found in Figures~\activityTimecourse,
\cardioTimecourse, and \sleepTimecourse.  \textbf{c. Timecourses of physical
  activity, weighted by mental health measures.}  The timecourses in
each panel display the average daily step counts (top panel) or
duration of low-intensity activity (bottom panel).  The colored lines
show average activity dynamics weighted by self-reported stress levels
at the start of the experiment (purple) and self-reported ``typical''
stress levels (pink).  The baseline curves (black) display the
average across all participants (re-plotted in Panel C to illustrate
scale differences across panels).  Timecourses for additional mental
health-related and fitness-related measures may be found in
Figures~\activityTimecourseMH, \cardioTimecourseMH, and \sleepTimecourseMH. Error ribbons in all panels denote
the standard error of the mean.  Horizontal lines below each panel's
timecourses denote intervals over which each weighted measure (color) differs
from the unweighted baseline (via a paired sample two-sided $t$-test
of the weighted mean values for each measure within a 30-day window around each timepoint; horizontal lines denote $p <
0.05$, corrected).}
\label{fig:dynamics}
\end{figure}

@@

One important limitation of our study is that we cannot distinguish
correlations between different measures from potential causal
effects.  For example, we cannot know (from our study) whether
engaging in particular forms of physical activity \textit{causes} changes in
memory performance or mental health, or whether (alternatively) people
who tend to engage in similar forms of physical activity also happen to
exhibit similar memory and/or mental health profiles.  In other words,
an overlapping set of processes or person-specific attributes may lead
someone to both form particular habits around physical activity and display high or low performance on a
given memory test.  We do not know whether memory performance or
aspects of mental health might be manipulated or influenced by
changing the patterns of physical activity someone engages in.  For
this reason, we have been careful to frame our findings as
correlations and associations, rather than to imply knowledge about a causal direction
of our findings.

>>>>

\paragraph{Languages.}
All participants reported that they were fluent in either 1 or 2
languages (25\textsuperscript{th} percentile: 1;
50\textsuperscript{th} percentile: 1; 75\textsuperscript{th}
percentile: 1), and that they were ``familiar'' with between 1 and 11
languages (25\textsuperscript{th} percentile: 1;
50\textsuperscript{th} percentile: 2; 75\textsuperscript{th}
percentile: 3).

@@

After studying all 10 flashcards, participants were given a multiple
choice memory test where they were shown a series of novel
photographs, each depicting one of the 10 words they had learned.
They were asked to select which (of 4 unique options) Gaelic word went
with the given picture.  The 3 incorrect options were selected at
random (with replacement across trials), and the orders in which the
choices appeared to the participant were also randomized.  Each of the
10 words they had learned was tested exactly once.

@@

\section*{Results}
Before testing our main hypotheses, we examined the behavioral data
from each of four memory tasks (Fig.~\ref{fig:tasks}): a random word list learning ``free
recall'' task; a naturalistic recall task whereby participants watched
a short video and then recounted the narrative; a foreign language
``flashcards'' task; and a spatial learning task.  Each of the first
three tasks (free recall, naturalistic recall, and the flashcards
task) included both an immediate (short term) memory test and a
delayed (long term) memory test.  The spatial learning task included
only an immediate test. Participants in all four tasks exhibited
general trends and tendencies that have been previously reported in
prior work.  We were also interested in characterizing the variability
in task performance across participants.  For example, if all
participants exhibited near-identical behaviors or performance on a
given task, we would be unable to identify how memory performance on
that task varied with mental health or physical activity.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/combined_correlations_main}
\caption{\textbf{Summaries of correlations between behavioral,
    fitness, and mental health measures.}  The reported values in the
  tables reflect correlations between each pair of measures.  Only statistically reliable correlations
  ($p < 0.05$, corrected) are displayed.  \textbf{a. Correlations
    between behavioral and mental health measures.}  We adjusted each
  task's behavioral measure(s) such that more positive values reflect
  better performance on the given task.  We used participants' mean
  recall accuracies to characterize performance on the free recall and
  foreign language flashcards tasks, and mean precisions to
  characterize performance on the naturalistic recall tasks.  We
  characterized performance on the spatial learning task using the
  (inverted and normalized) intercepts and slopes of linear
  regressions on mean estimation errors as a function of the numbers of
  studied shapes (also see Figs.~\ref{fig:immediate_behavior},
  \ref{fig:delayed_behavior},~\frDetail,~\natDetail,~\vocabDetail,
  and~\spatialDetail).  For each mental health measure, more positive
  values denote greater severity of the given measure.  Typical and
  current stress levels were measured by self report.  Mental health
  information was inferred using each participants' list of
  self-reported medications (see \textit{Methods}).  Positive
  correlations indicate that better performance on a given behavioral
  task is associated with more severe mental health phenotypes.
  \textbf{b. Correlations between fitness and mental health measures.}
  For each fitness measure, more positive values denote higher
  observed scores (i.e., higher resting heart rate, more minutes of
  activity or time spent in each heart rate zone, or greater heart
  rate variability).  The mental health measures in this panel were
  treated as in Panel a.  \textbf{c. Correlations between fitness and
    behavioral measures.}  Each measure reflected in this panel was
  treated as in Panels a and b.}
\label{fig:corr_summaries}
\end{figure}

We also found reliable correlations between participants' fitness and
demographic measures and their behaviors in different tasks
(Fig.~\ref{fig:corr_summaries}; for a complete report, see Fig.~\allCorr).
For example, recent low-to-moderate-intensity (``fat burn'')
cardiovascular activity was positively correlated with immediate
($r = 0.44, p = 0.001$) and delayed ($r = 0.38, p = 0.031$) recall
performance on the naturalistic memory task.  Recent sedentary
(``out-of-range'') cardiovascular activity was negatively correlated
with performance on the spatial learning task
($r = -0.31, p = 0.042$), whereas recent high intensity (``peak'')
activity was positively correlated with performance on the spatial
learning task ($r = 0.34, p = 0.0002$).  Mental health indicators,
such as self-reported stress levels and medications were also
associated with differences in memory
(Figs.~\ref{fig:corr_summaries}a, \allCorr).  For example,
self-reported stress levels at the time of test were negatively
correlated with performance on the delayed memory test for the foreign
language flashcards task ($r = -0.29, p = 0.038$), whereas
participants who were medicated for anxiety and depression tended to
perform slightly (but reliably) \textit{better} on the immediate
memory test for the foreign language flashcards task
($r = 0.11, p < 0.0001$).  Mental health indicators were also
correlated with several fitness measures
(Fig.~\ref{fig:corr_summaries}c).  For example, participants with
higher resting heart rates were less likely to be hypothyroid ($r =
-0.33, p < 0.0001$).  Participants who engaged in more low-intensity
(``light'') activity tended to be less anxious and depressed ($r =
-0.12, p = 0.03$), whereas participants who engaged in more
high-intensity activity tended to report higher levels of current ($r
= 0.15, p = 0.027$) and typical ($r = 0.21, p = 0.012$) stress.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/weighted_timecourse_summary}
\caption{\textbf{Dynamics of physical activity vary with memory
    performance and mental health measures.}  \textbf{a. Daily step counts.}  Each
  timecourse is weighted by either performance on immediate recall tests
  (left panel) or on delayed recall tests (right panel).  The black
  (baseline) timecourses display the (unweighted) average across all
  participants.  \textbf{b. Daily duration (in minutes) of low-intensity physical
    activity.}  Timecourses are displayed in the same format and color
scheme as those in Panel A.  Analogous timecourses for additional
fitness-related measures may be found in Figures~\activityTimecourse,
\cardioTimecourse, and \sleepTimecourse.  \textbf{c. Timecourses of physical
  activity, weighted by mental health measures.}  The timecourses in
each panel display the average daily step counts (top panel) or
duration of low-intensity activity (bottom panel).  The colored lines
show average activity dynamics weighted by self-reported stress levels
at the start of the experiment (purple) and self-reported ``typical''
stress levels (pink).  The baseline curves (black) display the
average across all participants (re-plotted in Panel C to illustrate
scale differences across panels).  Timecourses for additional mental
health-related and fitness-related measures may be found in
Figures~\activityTimecourseMH, \cardioTimecourseMH, and \sleepTimecourseMH. Error ribbons in all panels denote
the standard error of the mean.  Horizontal lines below each panel's
timecourses denote intervals over which each weighted measure (color) differs
from the unweighted baseline (via a paired sample two-sided $t$-test
of the weighted mean values for each measure within a 30-day window around each timepoint; horizontal lines denote $p <
0.05$, corrected).}
\label{fig:dynamics}
\end{figure}

@@

One important limitation of our study is that we cannot distinguish
correlations between different measures from potential causal
effects.  For example, we cannot know (from our study) whether
engaging in particular forms of physical activity \textit{causes} changes in
memory performance or mental health, or whether (alternatively) people
who tend to engage in similar forms of physical activity also happen to
exhibit similar memory and/or mental health profiles.  In other words,
an overlapping set of processes or person-specific attributes may lead
someone to both form particular habits around physical activity and display high or low performance on a
given memory test.  We do not know whether memory performance or
aspects of mental health might be manipulated or influenced by
changing the patterns of physical activity someone engages in.  For
this reason, we have been careful to frame our findings as
correlations and associations, rather than to imply knowledge about causal directions
of our findings.
